---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-07-6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE,
  fig.align = "center")
```


```{r}
library(tidyverse)
library(tidymodels)
library(patchwork)

```

```{r}
# Training Dataset

## Create binary outcome (BlueTarp)

haiti_train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')

haiti_train$BlueTarp <- ifelse(haiti_train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp' )

```
# Data Quality Checks

```{r}
# Normalize Color variables 

#haiti_train$red_scale <- haiti_train$Red/255 
#haiti_train$green_scale <- haiti_train$Green/255 
#haiti_train$blue_scale <- haiti_train$Blue/255
```

# Exploratory Data Analysis

## Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(haiti_train$BlueTarp)
```

## Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

haiti_train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Non-Tarp' = 'darkred','Blue Tarp' = 'skyblue'))
  
```

### Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(haiti_train, aes(x = BlueTarp, y = Red)) +
  geom_boxplot()
b2 <- ggplot(haiti_train, aes(x = BlueTarp, y = Green)) +
  geom_boxplot()
b3 <- ggplot(haiti_train, aes(x = BlueTarp, y = Blue)) +
  geom_boxplot()
b1+b2+b3
```

### Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(haiti_train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(haiti_train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(haiti_train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```


# Holdout Data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)

```

## Hold-Out Data: Unzip the file
```{r}
#Create a folder
unzip("HoldoutData.zip", exdir = "Holdout")

#Check files
list.files("Holdout")
```

## Parse single ENVI ROI file (Holdout Dataset Cleaning)

```{r}
parse_envi_roi <- function(filepath, label) {       #convert envi roi text file into the format I want
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df

```{r}

files <- tribble(
  ~path,                                      ~label,
  "orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  "orthovnir067_ROI_Blue_Tarps.txt",           1,
  "orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir069_ROI_Blue_Tarps.txt",           1,
  "orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir078_ROI_Blue_Tarps.txt",           1,
  "orthovnir078_ROI_NON_Blue_Tarps.txt",       0
)

haiti_holdout <- pmap_dfr(
 files,~ parse_envi_roi(file.path("Holdout", ..1), ..2))

cat("Total rows:", scales::comma(nrow(haiti_holdout)), "\n\n")
```

## Data quality checks

```{r}
cat("Dimensions \n")
 
cat("\n Missing values \n")
print(colSums(is.na(haiti_holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
haiti_holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(haiti_holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- haiti_holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
haiti_holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```

```{r}
df <- haiti_holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(df)

ggplot(df, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```

## KNN Analysis (Sienna)

```{r}
#install.packages("kknn")
library(kknn)

set.seed(100)

# make factor levels consistent
haiti_train <- haiti_train |>
  mutate(
    BlueTarp = factor(BlueTarp, levels = c("Non-Tarp", "Blue Tarp"))
  )

knn_rec <- recipe(BlueTarp ~ Red + Green + Blue, data = haiti_train) |>
  step_normalize(all_numeric_predictors())

# Cross-validation
folds <- vfold_cv(haiti_train, v = 10, strata = BlueTarp)

# Create KNN model
knn_spec <- nearest_neighbor(
  engine = "kknn",
  mode = "classification",
  neighbors = tune()
)

# Workflow
knn_wf <- workflow() |>
  add_recipe(knn_rec) |>
  add_model(knn_spec)

# Tuning range
knn_grid <- tibble(neighbors = 2:40)

# Metrics
knn_metrics <- metric_set(
  accuracy,
  precision,
  recall,
  f_meas,
  roc_auc,
  spec
)

# Tune K 
knn_results <- tune_grid(
  knn_wf,
  resamples = folds,
  grid = knn_grid,
  metrics = knn_metrics,
  control = control_grid(save_pred = TRUE)
)

# Plot tuning results
autoplot(knn_results) +
  theme_bw() +
  theme(
    text = element_text(size = 13)
  )

# Choose best K
show_best(knn_results, metric = "accuracy", n = 5)
show_best(knn_results, metric = "precision", n = 5)
show_best(knn_results, metric = "recall", n = 5)
show_best(knn_results, metric = "f_meas", n = 5)
show_best(knn_results, metric = "roc_auc", n = 5)

best_knn <- select_best(knn_results, metric = "roc_auc")
best_knn
```

```{r}
# selecting best nodel and fitting final model
final_knn_wf <- finalize_workflow(knn_wf, best_knn)

# Fit final KNN model
final_knn <- fit(final_knn_wf, data = haiti_train)
```


```{r}
exists("final_knn")
```

```{r}
class(final_knn)
```


```{r}
# Make holdout outcome factor levels match training data
haiti_holdout <- haiti_holdout |>
  mutate(
    BlueTarp = factor(BlueTarp, levels = levels(haiti_train$BlueTarp))
  )

# Predict on holdout data
knn_holdout_pred <- predict(final_knn, new_data = haiti_holdout, type = "class") |>
  bind_cols(predict(final_knn, new_data = haiti_holdout, type = "prob")) |>
  bind_cols(haiti_holdout |> select(BlueTarp)) |>
  mutate(
    .pred_class = factor(.pred_class, levels = levels(BlueTarp))
  )

names(knn_holdout_pred)

# Accuracy, precision, recall, F1, specificity
knn_holdout_pred |>
  metric_set(accuracy, precision, recall, f_meas, spec)(
    truth = BlueTarp,
    estimate = .pred_class,
    event_level = "second"
  )

# ROC AUC
knn_holdout_pred |>
  roc_auc(
    truth = BlueTarp,
    '.pred_Blue Tarp',
    event_level = "second"
  )

# Evaluate holdout performance
names(knn_holdout_pred)
  
```


Result interpretation: The number of neighbors (K) is a tunable hyperparameter in the K-Nearest Neighbors (KNN) algorithm. We evaluated values of K from 2 to 40 using 10-fold cross-validation on the training data. This tuning process identified the value of K that provided the best balance btween model complexity and predictive performance while helping to avoid overfitting with very small K values and underdefitting with very large K values. The optimal value selected was K=25, and this value was used to fit the final KNN model. 

The KNN results indicate that the RGB (Red, Green, and Blue) pixel intensity values are highly informative for distinguishing Blue Tarp from Non-Tarp pixels. Using 10-fold cross-validation, the model achieved an average ROC AUC of approximately 0.995, indicating excellent discriminative ability. In other words, the model is able to almost perfectly distinguish Blue Tarp pixels from Non-Tarp pixels. The overall accuracy was also very high at approximately 99.7%. However, because the dataset is highly imbalanced, accuracy alone is not an adequate measure of performance. Therefore, we also examined sensitivity (recall), specificity, precision, F1 score, and ROC AUC.

The tuning plots show that as K increases, the predictions become smoother and the model becomes less sensitive, resulting in a slight decrease in recall. We also observed that ROC AUC continued to increase slightly while the F1 score decreased. This behavior is expected because these metrics evaluate different aspects of model performance. ROC AUC measures how well the model ranks observations across all possible classification thresholds, whereas the F1 score evaluates the balance between precision and recall at a single decision threshold. Consequently, the two metrics do not necessarily reach their maximum values at the same value of K. Overall, the KNN model demonstrated outstanding predictive performance and showed that RGB color information alone is highly effective for identifying blue tarp pixels in this dataset.
