---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-06-08"
output: html_document
---
# Training Data
## Preprocess the Data

Create binary variable from Class

```{r}
haiti_train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')

haiti_train$BlueTarp <- ifelse(haiti_train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp' )
```


```{r}
# Normalize Color variables

# haiti_train$red_scale <- haiti_train$Red/255 
# haiti_train$green_scale <- haiti_train$Green/255 
# haiti_train$blue_scale <- haiti_train$Blue/255
```

## Exploratory Data Analysis

Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(haiti_train$BlueTarp)
```

Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

haiti_train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Blue Tarp' = 'skyblue', 'Non-Tarp' = 'darkred'))
  
```

Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(haiti_train, aes(x = BlueTarp, y = Red)) +
  geom_boxplot()
b2 <- ggplot(haiti_train, aes(x = BlueTarp, y = Green)) +
  geom_boxplot()
b3 <- ggplot(haiti_train, aes(x = BlueTarp, y = Blue)) +
  geom_boxplot()
b1+b2+b3
```

Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(haiti_train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(haiti_train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(haiti_train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```

# Holdout Data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
```


## Hold-Out Data: Unzip the file
```{r}
#Create a folder
unzip("HoldOutData.zip", exdir = "HoldOut")

#Check files
list.files("HoldOut")
```

## Parse single ENVI ROI file
```{r}
parse_envi_roi <- function(filepath, label) {
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df
```{r}
files <- tribble(
  ~path,                                      ~label,
  "orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  "orthovnir067_ROI_Blue_Tarps.txt",           1,
  "orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir069_ROI_Blue_Tarps.txt",           1,
  "orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir078_ROI_Blue_Tarps.txt",           1,
  "orthovnir078_ROI_NON_Blue_Tarps.txt",       0
)
 
holdout <- pmap_dfr(files, ~ parse_envi_roi(..1, ..2))
cat("Total rows:", scales::comma(nrow(holdout)), "\n\n")
```

## Data quality checks
```{r}
cat("Dimensions \n")
glimpse(holdout)
 
cat("\n Missing values \n")
print(colSums(is.na(holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```


```{r}
df <- holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(df)

ggplot(df, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```



Logistic Regression:

Base Logistic Regression:
```{r}
formula <- BlueTarp ~ Red + Green + Blue

# factor BlueTarp for both datasets
haiti_train <- haiti_train |>
  mutate(BlueTarp = factor(BlueTarp))
holdout <- holdout |>
  mutate(BlueTarp = factor(BlueTarp))

# make positive class for holdout set Blue Tarp to match training set
holdout$BlueTarp <- factor(holdout$BlueTarp, levels = levels(haiti_train$BlueTarp))

# create logistic regression model, fitting on the training data
logreg_model <- logistic_reg(mode = "classification", engine = "glm") |>
  fit(formula, haiti_train)
```

```{r}
# plot ROC curves for training and test data
get_roc_plot <- function(model, data, model_name) {
  predictions <- model |>
    augment(data)
  auc <- roc_auc(predictions, BlueTarp, `.pred_Blue Tarp`, event_level = "first")
  title <- paste0(model_name, " (AUC=", round(auc$.estimate, 3), ")")
  roc_data <- predictions |>
    roc_curve(truth = BlueTarp, `.pred_Blue Tarp`, event_level = "first")
  g <- autoplot(roc_data) +
    labs(title = title)
  return(g)
}

g1 <- get_roc_plot(logreg_model, holdout, "Logreg test")
g2 <- get_roc_plot(logreg_model, haiti_train, "Logreg train")

g1+g2
```
The ROC-AUC is similar between training and test data, so there is no evidence of overfitting.

```{r}
#| fig.cap: F-measure as a function of threshold for the three classification models
#| fig.alt: F-measure as a function of threshold for the three classification models
#| fig.width: 8
#| fig.height: 4
#| out.width: 100%

# determine threshold that maximizes f-measure
threshold_scan <- function(model, data, model_name) {
  threshold_data <- model |>
    augment(data) |>
    probably::threshold_perf(BlueTarp, `.pred_Blue Tarp`,
      thresholds = seq(0.05, 0.95, 0.01), event_level = "first",
      metrics = metric_set(f_meas))
  opt_threshold <- threshold_data |>
    arrange(-.estimate) |>
    first()
  g <- ggplot(threshold_data, aes(x = .threshold, y = .estimate)) +
    geom_line() +
    geom_point(data = opt_threshold, color = "red", size = 2) +
    labs(title = model_name, x = "Threshold", y = "F-Measure") +
    coord_cartesian(ylim = c(0.8, 1))
  return(list(
    graph = g,
    threshold = opt_threshold |>
      pull(.threshold)
  ))
}

logreg_threshold <- threshold_scan(logreg_model, haiti_train, "Logistic regression")
logreg_threshold
```
The optimal threshold is 0.21.

```{r}
custom_metrics <- metric_set(accuracy, precision, sens, spec, f_meas)

predict_at_threshold <- function(model, data, threshold) {
  return(
    model |>
      augment(data) |>
      mutate(.pred_class = make_two_class_pred(`.pred_Blue Tarp`,
          c("Blue Tarp", "Non-Tarp"), threshold = threshold),
          .pred_class = factor(.pred_class, levels=levels(BlueTarp))
      )
  )
}

calculate_metrics_at_threshold <- function(model, train, holdout,
  model_name, threshold) {
  df <- bind_rows(
    bind_cols(
      model = model_name, dataset = "holdout", threshold = threshold,
      custom_metrics(predict_at_threshold(model, holdout, threshold),
        truth = BlueTarp, estimate = .pred_class, event_level = "first"),
    ),
  )
  return(df)
}

calculate_metrics_at_threshold(logreg_model,haiti_train, holdout, "Logistic regression", logreg_threshold$threshold)|> 
  arrange(dataset)
```


10-fold Cross-Validation:
```{r}
set.seed(6030)

custom_metrics <- metric_set(accuracy, precision, sens, spec, f_meas, roc_auc)

# fit model using 10-fold cross-validation
logreg_rec <- recipe(formula, data = haiti_train)

logreg_spec <- logistic_reg(mode = "classification", engine = "glm")

logreg_wf <- workflow() |>
  add_recipe(logreg_rec) |>
  add_model(logreg_spec)

resamples <- vfold_cv(haiti_train, v = 10, strata = BlueTarp)
cv_control <- control_resamples(save_pred = TRUE)

logreg_cv <- fit_resamples(logreg_wf, resamples,
  metrics = custom_metrics,
  control = cv_control)

# report metrics of the model
collect_metrics(logreg_cv)
```

```{r}
# ROC-AUC curve for 10-fold cross-validation
collect_predictions(logreg_cv) |>
  roc_curve(truth = BlueTarp, `.pred_Blue Tarp`, event_level = "first") |>
  autoplot()
```

```{r}
# fit model on full training set and estimate performance metrics on test set
logreg_fit <- logreg_wf |> 
  fit(haiti_train)

augment(logreg_fit, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

Formatted like LDA and QDA:
```{r}
preds_logreg <- predict(logreg_fit, new_data = holdout)

prob_preds_logreg <- predict(logreg_fit, new_data = holdout, type = "prob")

results_logreg <- bind_cols(holdout, preds_logreg, prob_preds_logreg)

cm_logreg <- conf_mat(results_logreg, truth = BlueTarp, estimate = .pred_class)

acc_logreg <- accuracy(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_logreg <- precision(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_logreg <- recall(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_logreg <- f_meas(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

metrics_df <- tibble(
  model = character(),
  accuracy = numeric(),
  precision = numeric(),
  recall = numeric(),
  f1_score = numeric(),
  conf_matrix = list()
)

metrics_df <- metrics_df |>
  add_row(
    model = "Logistic Regression",
    accuracy = acc_logreg$.estimate,
    precision = prec_logreg$.estimate,
    recall = recall_logreg$.estimate,
    f1_score = f1_logreg$.estimate,
    conf_matrix = list(cm_logreg)
  )

metrics_df
```


Principle Components:
```{r}
# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = haiti_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pca_plot <- autoplot(tune_results)
pca_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(haiti_train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
# repeat above for PLS

# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = haiti_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pls(all_numeric_predictors(), outcome = 'BlueTarp', num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pls_plot <- autoplot(tune_results)
pls_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(haiti_train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

Results from PCA and PLS are the same as 10-fold cross-validation

```{r}
pca_plot + pls_plot
```
Using both PCA and PLS maintained all 3 components, so the metrics are identical to the 10-fold cross-validation.

