---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-07-08"
output: html_document
---
# Training Data
## Preprocess the Data

Create binary variable from Class

```{r}
train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')

train$BlueTarp <- ifelse(train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp' )
```


## Exploratory Data Analysis

Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(train$BlueTarp)
```

Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Blue Tarp' = 'skyblue', 'Non-Tarp' = 'darkred'))
  
```

Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(train, aes(x = BlueTarp, y = Red)) +
  geom_boxplot()
b2 <- ggplot(train, aes(x = BlueTarp, y = Green)) +
  geom_boxplot()
b3 <- ggplot(train, aes(x = BlueTarp, y = Blue)) +
  geom_boxplot()
b1+b2+b3
```

Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```

# Holdout Data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
```


## Hold-Out Data: Unzip the file
```{r}
#Create a folder
unzip("HoldOutData.zip", exdir = "HoldOut")

#Check files
list.files("HoldOut")
```

## Parse single ENVI ROI file
```{r}
parse_envi_roi <- function(filepath, label) {
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df
```{r}
files <- tribble(
  ~path,                                      ~label,
  "orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  "orthovnir067_ROI_Blue_Tarps.txt",           1,
  "orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir069_ROI_Blue_Tarps.txt",           1,
  "orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir078_ROI_Blue_Tarps.txt",           1,
  "orthovnir078_ROI_NON_Blue_Tarps.txt",       0
)
 
holdout <- pmap_dfr(files, ~ parse_envi_roi(..1, ..2))
cat("Total rows:", scales::comma(nrow(holdout)), "\n\n")
```

## Data quality checks
```{r}
cat("Dimensions \n")
glimpse(holdout)
 
cat("\n Missing values \n")
print(colSums(is.na(holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```


```{r}
df <- holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(df)

ggplot(df, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```



Logistic Regression:
```{r}
formula <- BlueTarp ~ Red + Green + Blue

# factor BlueTarp for both datasets
train <- train |>
  mutate(BlueTarp = factor(BlueTarp))
holdout <- holdout |>
  mutate(BlueTarp = factor(BlueTarp))

# make positive class for holdout set Blue Tarp to match training set
holdout$BlueTarp <- factor(holdout$BlueTarp, levels = levels(train$BlueTarp))

# create logistic regression model, fitting on the training data
logreg_model <- logistic_reg(mode = "classification", engine = "glm") |>
  fit(formula, train)
```


10-fold Cross-Validation:
```{r}
set.seed(123)

custom_metrics <- metric_set(accuracy, precision, sens, spec, f_meas, roc_auc)

# fit model using 10-fold cross-validation
rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors())

logreg_spec <- logistic_reg(mode = "classification", engine = "glm")

logreg_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(logreg_spec)

resamples <- vfold_cv(train, v = 10, strata = BlueTarp)
cv_control <- control_resamples(save_pred = TRUE)

logreg_cv <- fit_resamples(logreg_wf, resamples,
  metrics = custom_metrics,
  control = cv_control)

# report metrics of the model
collect_metrics(logreg_cv)
```

```{r}
# ROC-AUC curve for 10-fold cross-validation
collect_predictions(logreg_cv) |>
  roc_curve(truth = BlueTarp, `.pred_Blue Tarp`, event_level = "first") |>
  autoplot()
```

```{r}
# fit model on full training set and estimate performance metrics on test set
logreg_fit <- logreg_wf |> 
  fit(train)

augment(logreg_fit, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

Principle Components:
```{r}
# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pca_plot <- autoplot(tune_results)
pca_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
# repeat above for PLS

# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pls(all_numeric_predictors(), outcome = 'BlueTarp', num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pls_plot <- autoplot(tune_results)
pls_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
pca_plot + pls_plot
```

L1/L2 Regularization:
```{r}
logreg_reg_spec <- logistic_reg(engine = "glmnet", mode = "classification",
  penalty = tune(), mixture = tune())

linreg_reg_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(logreg_reg_spec)

parameters <- extract_parameter_set_dials(linreg_reg_wf)

model_bayes <- tune_bayes(linreg_reg_wf, resamples = resamples,
  param_info = parameters, iter = 25)

autoplot(model_bayes)
```

```{r}
show_best(model_bayes, metric = "roc_auc", n=1)

logreg_best_parameters <- select_best(model_bayes, metric = "roc_auc") |>
  select(-.config)

logreg_best_workflow <- linreg_reg_wf |>
  finalize_workflow(logreg_best_parameters) |>
  fit(train)

augment(logreg_best_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```


LDA:
```{r}
library(discrim)

# Omit Class due to inherent collinearity with BlueTarp
lda_model <- discrim_linear() |>
  set_engine("MASS") |>
  set_mode("classification")

lda_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(lda_model)

cv_results_lda <- fit_resamples(
  lda_wf,
  resamples = resamples,
  control = control_resamples(save_pred = TRUE)
)

lda_fit <- fit(lda_wf, data = train)

lda_engine <- extract_fit_engine(lda_fit)
```


```{r}
preds_lda <- predict(lda_fit, new_data = holdout)

prob_preds_lda <- predict(lda_fit, new_data = holdout, type = "prob")

results_lda <- bind_cols(holdout, preds_lda, prob_preds_lda)

cm_lda <- conf_mat(results_lda, truth = BlueTarp, estimate = .pred_class)

acc_lda <- accuracy(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_lda <- precision(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_lda <- recall(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_lda <- f_meas(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_lda <- roc_auc(
  results_lda,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)

metrics_df <- metrics_df |>
  add_row(
    model = "LDA",
    accuracy = acc_lda$.estimate,
    precision = prec_lda$.estimate,
    recall = recall_lda$.estimate,
    f1_score = f1_lda$.estimate,
    roc_auc = roc_auc_lda$.estimate,
    conf_matrix = list(cm_lda)
  )

metrics_df
```



QDA: 
```{r}
# Omit Class due to inherent collinearity with BlueTarp
qda_model <- discrim_quad() |>
  set_engine("MASS")

qda_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(qda_model)

cv_results_qda <- fit_resamples(
  qda_wf,
  resamples = resamples,
  control = control_resamples(save_pred = TRUE)
)

qda_fit <- fit(qda_wf, data = train)

qda_engine <- extract_fit_engine(qda_fit)
```

```{r}
preds_qda <- predict(qda_fit, new_data = holdout)

prob_preds_qda <- predict(qda_fit, new_data = holdout, type = "prob")

results_qda <- bind_cols(holdout, preds_qda, prob_preds_qda)

cm_qda <- conf_mat(results_qda, truth = BlueTarp, estimate = .pred_class)

acc_qda <- accuracy(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_qda <- precision(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_qda <- recall(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_qda <- f_meas(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_qda <- roc_auc(
  results_qda,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)


metrics_df <- metrics_df |>
  add_row(
    model = "QDA",
    accuracy = acc_qda$.estimate,
    precision = prec_qda$.estimate,
    recall = recall_qda$.estimate,
    f1_score = f1_qda$.estimate,
    roc_auc = roc_auc_qda$.estimate,
    conf_matrix = list(cm_qda)
  )

metrics_df
```


KNN:
```{r}
#install.packages("kknn")
library(kknn)

set.seed(100)

# make factor levels consistent
haiti_train <- haiti_train |>
  mutate(
    BlueTarp = factor(BlueTarp, levels = c("Non-Tarp", "Blue Tarp"))
  )

knn_rec <- recipe(BlueTarp ~ Red + Green + Blue, data = haiti_train) |>
  step_normalize(all_numeric_predictors())

# Cross-validation
folds <- vfold_cv(haiti_train, v = 10, strata = BlueTarp)

# Create KNN model
knn_spec <- nearest_neighbor(
  engine = "kknn",
  mode = "classification",
  neighbors = tune()
)

# Workflow
knn_wf <- workflow() |>
  add_recipe(knn_rec) |>
  add_model(knn_spec)

# Tuning range
knn_grid <- tibble(neighbors = 2:40)

# Metrics
knn_metrics <- metric_set(
  accuracy,
  precision,
  recall,
  f_meas,
  roc_auc,
  spec
)

# Tune K 
knn_results <- tune_grid(
  knn_wf,
  resamples = folds,
  grid = knn_grid,
  metrics = knn_metrics,
  control = control_grid(save_pred = TRUE)
)

# Plot tuning results
autoplot(knn_results) +
  theme_bw() +
  theme(
    text = element_text(size = 13)
  )

# Choose best K
show_best(knn_results, metric = "accuracy", n = 5)
show_best(knn_results, metric = "precision", n = 5)
show_best(knn_results, metric = "recall", n = 5)
show_best(knn_results, metric = "f_meas", n = 5)
show_best(knn_results, metric = "roc_auc", n = 5)

best_knn <- select_best(knn_results, metric = "roc_auc")
best_knn
```

```{r}
# selecting best nodel and fitting final model
final_knn_wf <- finalize_workflow(knn_wf, best_knn)

# Fit final KNN model
final_knn <- fit(final_knn_wf, data = haiti_train)
```


```{r}
exists("final_knn")
```

```{r}
class(final_knn)
```


```{r}
# Make holdout outcome factor levels match training data
haiti_holdout <- haiti_holdout |>
  mutate(
    BlueTarp = factor(BlueTarp, levels = levels(haiti_train$BlueTarp))
  )

# Predict on holdout data
knn_holdout_pred <- predict(final_knn, new_data = haiti_holdout, type = "class") |>
  bind_cols(predict(final_knn, new_data = haiti_holdout, type = "prob")) |>
  bind_cols(haiti_holdout |> select(BlueTarp)) |>
  mutate(
    .pred_class = factor(.pred_class, levels = levels(BlueTarp))
  )

names(knn_holdout_pred)

# Accuracy, precision, recall, F1, specificity
knn_holdout_pred |>
  metric_set(accuracy, precision, recall, f_meas, spec)(
    truth = BlueTarp,
    estimate = .pred_class,
    event_level = "second"
  )

# ROC AUC
knn_holdout_pred |>
  roc_auc(
    truth = BlueTarp,
    '.pred_Blue Tarp',
    event_level = "second"
  )

# Evaluate holdout performance
names(knn_holdout_pred)
  
```


Random Forests:
```{r}
# define model
rf_spec <- rand_forest(mode = "classification", trees = 1000, min_n = tune(), 
  mtry = tune()) |>
    set_engine("ranger", importance = "impurity")

# create workflow
rf_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)

# tuning
rf_grid <- grid_regular(mtry(range = c(1,3)),
  min_n(range = c(2, 30)),
  levels = 5)

tuned_rf <- tune_grid(rf_wf, resamples = resamples,
  grid = rf_grid, metrics = cv_metrics)
```

```{r}
autoplot(tuned_rf)
```
Performance is relatively stable for each metric across min_n

```{r}
# selecting best model and fitting final model

best_rf <- select_best(tuned_rf, metric = "roc_auc")
final_rf <- finalize_workflow(rf_wf, best_rf)

rf_fit <- fit(final_rf, data = train)
```

```{r}
preds_rf <- predict(rf_fit, new_data = holdout)

prob_preds_rf <- predict(rf_fit, new_data = holdout, type = "prob")

results_rf <- bind_cols(holdout, preds_rf, prob_preds_rf)

cm_rf <- conf_mat(results_rf, truth = BlueTarp, estimate = .pred_class)

acc_rf <- accuracy(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_rf <- precision(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_rf <- recall(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_rf <- f_meas(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

metrics_df <- metrics_df |>
  add_row(
    model = "Random Forests",
    accuracy = acc_rf$.estimate,
    precision = prec_rf$.estimate,
    recall = recall_rf$.estimate,
    f1_score = f1_rf$.estimate,
    conf_matrix = list(cm_rf)
  )

metrics_df
```

