---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-08-03"
output: html_document
---
# Training Data
## Preprocess the Data

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
```

Create binary variable from Class

```{r}
train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')

train$BlueTarp <- ifelse(train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp' )
```


## Exploratory Data Analysis

Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(train$BlueTarp)
```

Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Blue Tarp' = 'skyblue', 'Non-Tarp' = 'darkred'))
  
```

Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(train, aes(x = BlueTarp, y = Red, fill = BlueTarp)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
b2 <- ggplot(train, aes(x = BlueTarp, y = Green, fill = BlueTarp)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
b3 <- ggplot(train, aes(x = BlueTarp, y = Blue, fill = BlueTarp)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
b1+b2+b3
```

```{r}
haiti_long <- train |>
  pivot_longer(cols = c(Red, Blue, Green),
               names_to = "Channel",
               values_to = "Value")

ggplot(haiti_long,
       aes(Value, fill = factor(BlueTarp), color = factor(BlueTarp))) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Channel, ncol = 1) +
  labs(fill = "Blue Tarp", color = "Blue Tarp") +
  theme_bw()
```

Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```

Pairs plot
```{r}
library(GGally)

ggpairs(
  train,
  columns = c("Red", "Green", "Blue", "BlueTarp"),
  aes(color = BlueTarp, fill = BlueTarp, alpha = 0.5),
  upper = list(continuous = wrap("cor", size = 3), combo = "box_no_facet"),
  lower = list(continuous = wrap("points", size = 0.5, alpha = 0.3), combo = "box_no_facet")
)
```

```{r}
library(Rtsne)

set.seed(1234)
tsne_df <- train |>
  select(-BlueTarp) |>
  select(-Class)

tsne_tarp <- Rtsne(scale(tsne_df), perplexity = 40,
  pca = TRUE, theta = 0.5, check_duplicates = FALSE)

df <- data.frame(
  TSNE1 = tsne_tarp$Y[, 1],
  TSNE2 = tsne_tarp$Y[, 2],
  Red = train$Red,
  Green = train$Green,
  Blue = train$Blue
) |>
  mutate(pixel_color = rgb(Red, Green, Blue, maxColorValue = 255))

ggplot(df, aes(x = TSNE1, y = TSNE2, colour = pixel_color)) +
  geom_point() +
  scale_colour_identity() +
  labs(title = "t-SNE Plot Colored by Pixel RGB")
```

# Holdout Data

## Hold-Out Data: Unzip the file
```{r}
#Create a folder
unzip("HoldOutData.zip", exdir = "HoldOut")

#Check files
list.files("HoldOut")
```

## Parse single ENVI ROI file
```{r}
parse_envi_roi <- function(filepath, label) {
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df
```{r}
files <- tribble(
  ~path,                                      ~label,
  "HoldOut/orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  "HoldOut/orthovnir067_ROI_Blue_Tarps.txt",           1,
  "HoldOut/orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  "HoldOut/orthovnir069_ROI_Blue_Tarps.txt",           1,
  "HoldOut/orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  "HoldOut/orthovnir078_ROI_Blue_Tarps.txt",           1,
  "HoldOut/orthovnir078_ROI_NON_Blue_Tarps.txt",       0
)
 
holdout <- pmap_dfr(files, ~ parse_envi_roi(..1, ..2))
cat("Total rows:", scales::comma(nrow(holdout)), "\n\n")
```

## Data quality checks
```{r}
cat("Dimensions \n")
glimpse(holdout)
 
cat("\n Missing values \n")
print(colSums(is.na(holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```


```{r}
df <- holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(df)

ggplot(df, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```

Distribution comparison of training set with holdout set
```{r}
train_filtered <- train |>
  filter(BlueTarp == "Blue Tarp")
holdout_filtered <- holdout |>
  filter(BlueTarp == "Blue Tarp")

g1 <-ggplot(train_filtered, aes(Red)) +
  geom_histogram(fill='darkred', binwidth = 5)
g2 <- ggplot(train_filtered, aes(Green)) +
  geom_histogram(fill='lightgreen', binwidth = 5)
g3 <- ggplot(train_filtered, aes(Blue)) +
  geom_histogram(fill='skyblue', binwidth = 5)
g4 <-ggplot(holdout_filtered, aes(Red)) +
  geom_histogram(fill='darkred', binwidth = 5)
g5 <- ggplot(holdout_filtered, aes(Green)) +
  geom_histogram(fill='lightgreen', binwidth = 5)
g6 <- ggplot(holdout_filtered, aes(Blue)) +
  geom_histogram(fill='skyblue', binwidth = 5)

(g1+g2+g3) / (g4+g5+g6) +
  plot_annotation(title = "Train vs Holdout Distribution Comparison for Blue Tarp Class")
```

Logistic Regression:
```{r}
formula <- BlueTarp ~ Red + Green + Blue

# factor BlueTarp for both datasets
train <- train |>
  mutate(BlueTarp = factor(BlueTarp))
holdout <- holdout |>
  mutate(BlueTarp = factor(BlueTarp))

# make positive class for holdout set Blue Tarp to match training set
holdout$BlueTarp <- factor(holdout$BlueTarp, levels = levels(train$BlueTarp))

# create logistic regression model, fitting on the training data
logreg_model <- logistic_reg(mode = "classification", engine = "glm") |>
  fit(formula, train)
```


10-fold Cross-Validation:
```{r}
set.seed(123)

custom_metrics <- metric_set(accuracy, precision, sens, f_meas, roc_auc)

# fit model using 10-fold cross-validation
rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors())

logreg_spec <- logistic_reg(mode = "classification", engine = "glm")

logreg_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(logreg_spec)

resamples <- vfold_cv(train, v = 10, strata = BlueTarp)
cv_control <- control_resamples(save_pred = TRUE)

logreg_cv <- fit_resamples(logreg_wf, resamples,
  metrics = custom_metrics,
  control = cv_control)

# report metrics of the model
collect_metrics(logreg_cv)
```

```{r}
# ROC-AUC curve for 10-fold cross-validation
collect_predictions(logreg_cv) |>
  roc_curve(truth = BlueTarp, `.pred_Blue Tarp`, event_level = "first") |>
  autoplot()
```

```{r}
# fit model on full training set and estimate performance metrics on test set
logreg_fit <- logreg_wf |> 
  fit(train)

augment(logreg_fit, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
preds_logreg <- predict(logreg_fit, new_data = holdout)

prob_preds_logreg <- predict(logreg_fit, new_data = holdout, type = "prob")

results_logreg <- bind_cols(holdout, preds_logreg, prob_preds_logreg)

cm_logreg <- conf_mat(results_logreg, truth = BlueTarp, estimate = .pred_class)

acc_logreg <- accuracy(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_logreg <- precision(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_logreg <- recall(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_logreg <- f_meas(
  results_logreg,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_logreg <- roc_auc(
  results_logreg,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)

metrics_df <- tibble(
  model = character(),
  accuracy = numeric(),
  precision = numeric(),
  recall = numeric(),
  f1_score = numeric(),
  roc_auc = numeric(),
  conf_matrix = list()
)

metrics_df <- metrics_df |>
  add_row(
    model = "Logistic Regression",
    accuracy = acc_logreg$.estimate,
    precision = prec_logreg$.estimate,
    recall = recall_logreg$.estimate,
    f1_score = f1_logreg$.estimate,
    roc_auc = roc_auc_logreg$.estimate,
    conf_matrix = list(cm_logreg)
  )

metrics_df
```

Principle Components:
```{r}
# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pca_plot <- autoplot(tune_results)
pca_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
# repeat above for PLS

# tune to find best number of components
logreg_tune_rec <- recipe(formula, data = train) |>
  step_normalize(all_numeric_predictors()) |>
  step_pls(all_numeric_predictors(), outcome = 'BlueTarp', num_comp = tune())

logreg_tune_wf <- workflow() |>
  add_recipe(logreg_tune_rec) |>
  add_model(logreg_spec)

tune_results <- tune_grid(logreg_tune_wf,
  resamples = resamples, grid = grid_regular(num_comp(range = c(1,3)), levels = 3))

# plot tuning results
pls_plot <- autoplot(tune_results)
pls_plot

# select best parameters using roc-auc
best_parameters <- select_best(tune_results, metric = "roc_auc")

show_best(tune_results, n=1)

best_logreg_workflow <- logreg_tune_wf |>
  finalize_workflow(best_parameters) |>
  fit(train)

# compute metrics
augment(best_logreg_workflow, new_data = holdout) |>
  custom_metrics(truth = BlueTarp, estimate = .pred_class, `.pred_Blue Tarp`, event_level = "first")|>
  mutate(data = "Holdout")
```

```{r}
pca_plot + pls_plot
```

L1/L2 Regularization:
```{r}
logreg_reg_spec <- logistic_reg(engine = "glmnet", mode = "classification",
  penalty = tune(), mixture = tune())

linreg_reg_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(logreg_reg_spec)

parameters <- extract_parameter_set_dials(linreg_reg_wf)

model_bayes <- tune_bayes(linreg_reg_wf, resamples = resamples,
  param_info = parameters, iter = 25)

autoplot(model_bayes)
```

```{r}
show_best(model_bayes, metric = "roc_auc", n=1)

logreg_best_parameters <- select_best(model_bayes, metric = "roc_auc") |>
  select(-.config)

logreg_best_workflow <- linreg_reg_wf |>
  finalize_workflow(logreg_best_parameters) |>
  fit(train)
```

```{r}
preds_logreg_best <- predict(logreg_best_workflow, new_data = holdout)

prob_preds_logreg_best <- predict(logreg_best_workflow, new_data = holdout, type = "prob")

results_logreg_best <- bind_cols(holdout, preds_logreg_best, prob_preds_logreg_best)

cm_logreg_best <- conf_mat(results_logreg_best, truth = BlueTarp, estimate = .pred_class)

acc_logreg_best <- accuracy(
  results_logreg_best,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_logreg_best <- precision(
  results_logreg_best,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_logreg_best <- recall(
  results_logreg_best,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_logreg_best <- f_meas(
  results_logreg_best,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_logreg_best <- roc_auc(
  results_logreg_best,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)

metrics_df <- metrics_df |>
  add_row(
    model = "Logistic Regression Regularization",
    accuracy = acc_logreg_best$.estimate,
    precision = prec_logreg_best$.estimate,
    recall = recall_logreg_best$.estimate,
    f1_score = f1_logreg_best$.estimate,
    roc_auc = roc_auc_logreg_best$.estimate,
    conf_matrix = list(cm_logreg_best)
  )

metrics_df
```


LDA:
```{r}
library(discrim)

# Omit Class due to inherent collinearity with BlueTarp
lda_model <- discrim_linear() |>
  set_engine("MASS") |>
  set_mode("classification")

lda_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(lda_model)

cv_results_lda <- fit_resamples(
  lda_wf,
  resamples = resamples,
  control = control_resamples(save_pred = TRUE)
)

lda_fit <- fit(lda_wf, data = train)

lda_engine <- extract_fit_engine(lda_fit)
```

```{r}
preds_lda <- predict(lda_fit, new_data = holdout)

prob_preds_lda <- predict(lda_fit, new_data = holdout, type = "prob")

results_lda <- bind_cols(holdout, preds_lda, prob_preds_lda)

cm_lda <- conf_mat(results_lda, truth = BlueTarp, estimate = .pred_class)

acc_lda <- accuracy(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_lda <- precision(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_lda <- recall(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_lda <- f_meas(
  results_lda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_lda <- roc_auc(
  results_lda,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)

metrics_df <- metrics_df |>
  add_row(
    model = "LDA",
    accuracy = acc_lda$.estimate,
    precision = prec_lda$.estimate,
    recall = recall_lda$.estimate,
    f1_score = f1_lda$.estimate,
    roc_auc = roc_auc_lda$.estimate,
    conf_matrix = list(cm_lda)
  )

metrics_df
```

```{r}
lda_pred <- predict(lda_engine, newdata = train)

lda_df <- data.frame(
  LD1 = lda_pred$x[,1],
  BlueTarp = train$BlueTarp
)

library(ggplot2)

ggplot(lda_df, aes(x = LD1, fill = BlueTarp)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "LDA Discriminant Scores",
    x = "LD1",
    y = "Density"
  )
```

```{r}
coef_df <- data.frame(
  Variable = rownames(lda_engine$scaling),
  Coefficient = lda_engine$scaling[,1]
)

ggplot(coef_df,
       aes(x = reorder(Variable, Coefficient),
           y = Coefficient)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "LDA Feature Importance",
    x = "",
    y = "Coefficient"
  )
```

```{r}
library(GGally)

ggpairs(
    train,
    columns = c("Red","Green","Blue"),
    mapping = aes(color = BlueTarp)
)
```

```{r}
results_lda |>
    roc_curve(
        truth = BlueTarp,
        `.pred_Blue Tarp`,
        event_level="first"
    ) |>
    autoplot()
```

QDA: 
```{r}
qda_rec <- recipe(formula, data = train)
# Omit Class due to inherent collinearity with BlueTarp
qda_model <- discrim_quad() |>
  set_engine("MASS")

qda_wf <- workflow() |>
  add_recipe(qda_rec) |>
  add_model(qda_model)

cv_results_qda <- fit_resamples(
  qda_wf,
  resamples = resamples,
  control = control_resamples(save_pred = TRUE)
)

qda_fit <- fit(qda_wf, data = train)

qda_engine <- extract_fit_engine(qda_fit)
```

```{r}
preds_qda <- predict(qda_fit, new_data = holdout)

prob_preds_qda <- predict(qda_fit, new_data = holdout, type = "prob")

results_qda <- bind_cols(holdout, preds_qda, prob_preds_qda)

cm_qda <- conf_mat(results_qda, truth = BlueTarp, estimate = .pred_class)

acc_qda <- accuracy(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_qda <- precision(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_qda <- recall(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_qda <- f_meas(
  results_qda,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

roc_auc_qda <- roc_auc(
  results_qda,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)


metrics_df <- metrics_df |>
  add_row(
    model = "QDA",
    accuracy = acc_qda$.estimate,
    precision = prec_qda$.estimate,
    recall = recall_qda$.estimate,
    f1_score = f1_qda$.estimate,
    roc_auc = roc_auc_qda$.estimate,
    conf_matrix = list(cm_qda)
  )

metrics_df
```

```{r}
results_qda |>
    roc_curve(
        truth = BlueTarp,
        `.pred_Blue Tarp`,
        event_level = "first"
    ) |>
    autoplot()
```

```{r}
results_qda$Correct <-
    results_qda$BlueTarp ==
    results_qda$.pred_class

ggplot(results_qda,
       aes(Red,
           Blue,
           color = Correct)) +
    geom_point(alpha = .5) +
    theme_minimal() +
    labs(
      title = "QDA Classification Errors"
    )
```

```{r}
library(MASS)
library(plotly)
library(dplyr)

# Extract LDA coefficients
lda_coef <- lda_engine$scaling[,1]

# Calculate intercept
lda_intercept <- -sum(
  lda_engine$means[1,] * lda_coef
)


# Create Red-Green grid
plane_grid <- expand.grid(
  Red = seq(min(train$Red), max(train$Red), length.out = 60),
  Green = seq(min(train$Green), max(train$Green), length.out = 60)
)

plane_grid$Blue <- -(
  lda_intercept +
    lda_coef["Red"] * plane_grid$Red +
    lda_coef["Green"] * plane_grid$Green
) / lda_coef["Blue"]


# Keep plane inside observed RGB range
plane_grid <- plane_grid %>%
  filter(
    Blue >= min(train$Blue),
    Blue <= max(train$Blue)
  )


red_vals <- sort(unique(plane_grid$Red))
green_vals <- sort(unique(plane_grid$Green))


lda_plane_z <- matrix(
  plane_grid$Blue,
  nrow = length(red_vals),
  ncol = length(green_vals)
)

blue_tarp_points <- train %>%
  filter(BlueTarp == "Blue Tarp")


non_blue_tarp_points <- train %>%
  filter(BlueTarp != "Blue Tarp")

lda_3d_plot <- plot_ly() %>%

  # Blue tarp pixels
  add_trace(
    data = blue_tarp_points,
    x = ~Red,
    y = ~Green,
    z = ~Blue,
    type = "scatter3d",
    mode = "markers",
    marker = list(
      size = 3,
      color = "blue",
      opacity = 0.75
    ),
    name = "Blue Tarp",
    showlegend = TRUE
  ) %>%

  # Non-blue tarp pixels
  add_trace(
    data = non_blue_tarp_points,
    x = ~Red,
    y = ~Green,
    z = ~Blue,
    type = "scatter3d",
    mode = "markers",
    marker = list(
      size = 3,
      color = "gray",
      opacity = 0.5
    ),
    name = "Non-Blue Tarp",
    showlegend = TRUE
  ) %>%

  # LDA boundary plane
  add_surface(
    x = red_vals,
    y = green_vals,
    z = lda_plane_z,
    opacity = 0.55,
    colorscale = list(
      c(0,1),
      c("red","red")
    ),
    showscale = FALSE,
    name = "LDA Decision Boundary",
    showlegend = TRUE
  ) %>%

  layout(
    title = "LDA Decision Boundary in RGB Space",
    showlegend = TRUE,

    legend = list(
      title = list(
        text = "Legend"
      )
    ),

    scene = list(
      xaxis = list(
        title = "Red Intensity"
      ),
      yaxis = list(
        title = "Green Intensity"
      ),
      zaxis = list(
        title = "Blue Intensity"
      )
    )
  )


lda_3d_plot
```

```{r}
nrow(lda_boundary)
```

```{r}
library(patchwork)

lda_rb_plot
```

KNN:
```{r}
#install.packages("kknn")
library(kknn)
```

```{r}
cores <- max(1, parallel::detectCores() - 1)

cl <- parallel::makePSOCKcluster(cores)

doParallel::registerDoParallel(cl)
```

```{r}
# make factor levels consistent
set.seed(100)

# Create KNN model
knn_spec <- nearest_neighbor(
  engine = "kknn",
  mode = "classification",
  neighbors = tune()
)

# Workflow
knn_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(knn_spec)

# Tuning range
knn_grid <- tibble(neighbors = 2:40)

# Tune K 
knn_results <- tune_grid(
  knn_wf,
  resamples = resamples,
  grid = knn_grid,
  metrics = custom_metrics,
  control = control_grid(save_pred = TRUE)
)

# Plot tuning results
autoplot(knn_results) +
  theme_bw()+
  theme(
    text = element_text(size = 13)
  )

# Choose best K
show_best(knn_results, metric = "accuracy", n = 5)
show_best(knn_results, metric = "precision", n = 5)
show_best(knn_results, metric = "sens", n = 5)
show_best(knn_results, metric = "f_meas", n = 5)
show_best(knn_results, metric = "roc_auc", n = 5)

best_knn <- select_best(knn_results, metric = "roc_auc")
best_knn
```
```{r}
# selecting best model and fitting final model
final_knn_wf <- finalize_workflow(knn_wf, best_knn)

# Fit final KNN model
final_knn <- fit(final_knn_wf, data = train)
```


```{r}
# Predict on holdout data
knn_holdout_pred <- predict(final_knn, new_data = holdout, type = "class") |>
  bind_cols(predict(final_knn, new_data = holdout, type = "prob")) |>
  bind_cols(holdout |> select(BlueTarp) 
  )

names(knn_holdout_pred)

# Accuracy
acc_knn <- accuracy(
  knn_holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class
)

# Precision
prec_knn <- precision(
  knn_holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

# Recall / sensitivity
sens_knn <- recall(
  knn_holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

# F1 score
f1_knn <- f_meas(
  knn_holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

# ROC AUC — use exact probability-column name with backticks
auc_knn <- roc_auc(
  knn_holdout_pred,
  truth = BlueTarp,
  `.pred_Blue Tarp`,
  event_level = "first"
)

# Confusion matrix
cm_knn <- conf_mat(
  knn_holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class
)
```

```{r}
metrics_df <- metrics_df |>
  add_row(
    model = "KNN",
    accuracy = acc_knn$.estimate,
    precision = prec_knn$.estimate,
    recall = sens_knn$.estimate,
    f1_score = f1_knn$.estimate,
    roc_auc = auc_knn$.estimate,
    conf_matrix = list(cm_knn)
  )

metrics_df
```

```{r}
library(dplyr)
library(ggplot2)

# Extract ROC AUC results for each value of K
knn_auc_results <- knn_results |>
  collect_metrics() |>
  filter(.metric == "roc_auc")

# Identify the best K based on ROC AUC
best_auc_point <- knn_auc_results |>
  slice_max(order_by = mean, n = 1, with_ties = FALSE)

# Plot ROC AUC versus number of neighbors (K Plot)
ggplot(
  knn_auc_results,
  aes(x = neighbors, y = mean)
) +
  geom_line(linewidth = 1) +          #Error bars across resamples
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = mean - std_err,
      ymax = mean + std_err
    ),
    width = 0.35,
    color = "lightblue3",
    alpha = 0.7
  ) +
  # ROC AUC trend
  geom_line(
    color = "steelblue4",
    linewidth = 1.2
  ) +
  geom_point(
    color = "steelblue4",
    size = 2.5
  ) +

  # Highlight best K
  geom_vline(
    xintercept = best_auc_point$neighbors,
    color = "firebrick3",
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_point(
    data = best_auc_point,
    color = "firebrick3",
    size = 4.5
  ) +
    # Label best K
  annotate(
    "label",
    x = best_auc_point$neighbors,
    y = best_auc_point$mean,
    label = paste0(
      "Best K = ", best_auc_point$neighbors,
      "\nROC AUC = ", round(best_auc_point$mean, 4)
    ),
    color = "firebrick4",
    fill = "white",
    label.size = 0.3,
    vjust = -1.2,
    hjust = -0.05,
    size = 4
  ) +

  # Axis formatting
  scale_x_continuous(
    breaks = seq(2, 40, by = 2)
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.001),
    expand = expansion(mult = c(0.05, 0.15))
  ) +

  # Titles and labels
  labs(
    title = "KNN Tuning: ROC AUC vs. Number of Neighbors",
    subtitle = "Mean ROC AUC from 10-fold cross-validation",
    x = "Number of Neighbors (K)",
    y = "Mean ROC AUC",
    caption = "Error bars represent ±1 standard error across resamples."
  ) +
  # Presentation formatting
  theme_bw() +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(
      size = 17,
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 13,
      hjust = 0.5
    ),
    axis.title = element_text(
      size = 14,
      face = "bold"
    ),
    axis.text = element_text(size = 11),
    plot.caption = element_text(
      size = 10,
      hjust = 0
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(
      color = "grey90",
      linewidth = 0.4
    )
  )
```



```{r}
parallel::stopCluster(cl)
```


Random Forests:
```{r}
# define model
rf_spec <- rand_forest(mode = "classification", trees = 1000, min_n = tune(), 
  mtry = tune()) |>
    set_engine("ranger", importance = "impurity")

# create workflow
rf_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_spec)

# tuning
rf_grid <- grid_regular(mtry(range = c(1,3)),
  min_n(range = c(2, 30)),
  levels = 5)

tuned_rf <- tune_grid(rf_wf, resamples = resamples,
  grid = rf_grid, metrics = custom_metrics)
```

```{r}
autoplot(tuned_rf)
```
Performance is relatively stable for each metric across min_n

```{r}
# selecting best model and fitting final model

best_rf <- select_best(tuned_rf, metric = "roc_auc")
final_rf <- finalize_workflow(rf_wf, best_rf)

rf_fit <- fit(final_rf, data = train)
```

```{r}
preds_rf <- predict(rf_fit, new_data = holdout)

prob_preds_rf <- predict(rf_fit, new_data = holdout, type = "prob")

results_rf <- bind_cols(holdout, preds_rf, prob_preds_rf)

cm_rf <- conf_mat(results_rf, truth = BlueTarp, estimate = .pred_class)

acc_rf <- accuracy(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
)

prec_rf <- precision(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

recall_rf <- recall(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

f1_rf <- f_meas(
  results_rf,
  truth = BlueTarp,
  estimate = .pred_class,
  event_level = "first"
)

metrics_df <- metrics_df |>
  add_row(
    model = "Random Forests",
    accuracy = acc_rf$.estimate,
    precision = prec_rf$.estimate,
    recall = recall_rf$.estimate,
    f1_score = f1_rf$.estimate,
    conf_matrix = list(cm_rf)
  )

metrics_df
```

