---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-06-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE,
  fig.align = "center")
```

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
```

# Training Dataset

## Create binary outcome (BlueTarp)

```{r}
haiti_train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')
dim(haiti_train)
colSums(is.na(haiti_train))

haiti_train$BlueTarp <- ifelse(haiti_train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp' )
```

# Holdout Dataset <Data:Unzip> 

```{r}
#Create a folder
unzip("HoldOutData.zip", exdir = "HoldOut")

#Check files
list.files("HoldOut")
```
# Data Quality Checks

```{r}
# Normalize Color variables 

#haiti_train$red_scale <- haiti_train$Red/255 
#haiti_train$green_scale <- haiti_train$Green/255 
#haiti_train$blue_scale <- haiti_train$Blue/255
```

# Exploratory Data Analysis

## Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(haiti_train$BlueTarp)
```

## Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

haiti_train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Blue Tarp' = 'skyblue', 'Non-Tarp' = 'darkred'))
  
```

### Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(haiti_train, aes(x = BlueTarp, y = Red)) +
  geom_boxplot()
b2 <- ggplot(haiti_train, aes(x = BlueTarp, y = Green)) +
  geom_boxplot()
b3 <- ggplot(haiti_train, aes(x = BlueTarp, y = Blue)) +
  geom_boxplot()
b1+b2+b3
```

### Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(haiti_train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(haiti_train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(haiti_train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```

## Parse single ENVI ROI file (Holdout Dataset Cleaning)

```{r}
parse_envi_roi <- function(filepath, label) {       #convert envi roi text file into the format I want
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df

```{r}
## Load all files and combine into one df

#files <- tribble(
  #~path,                                      ~label,
 # "orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  #"orthovnir067_ROI_Blue_Tarps.txt",           1,
  #"orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  #"orthovnir069_ROI_Blue_Tarps.txt",           1,
  #"orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  #"orthovnir078_ROI_Blue_Tarps.txt",           1,
  #"orthovnir078_ROI_NON_Blue_Tarps.txt",       0
#)

#holdout_real <- pmap_dfr(
 # files,
  #~ parse_envi_roi(file.path("HoldOut", ..1), ..2)
#)

#cat("Total rows:", scales::comma(nrow(holdout_real)), "\n\n")
```

## Data quality checks

```{r}
cat("Dimensions \n")
glimpse(holdout)
 
cat("\n Missing values \n")
print(colSums(is.na(holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```

```{r}
df <- holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(df)

ggplot(df, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```

## KNN Analysis (Sienna)

```{r}
library(tidyverse)
library(tidymodels)
library(patchwork)

#install.packages("kknn")
library(kknn)

# Prepare data
# Split train/test

set.seed(1234)

split <- initial_split(haiti_train, prop = 0.8, strata = BlueTarp)

train <- training(split)
test <- testing(split)


# Preprocess
formula <- BlueTarp ~ Red + Green + Blue

rec <- recipe(formula, data=train) |>
  step_normalize(all_numeric_predictors())

# Cross-validation
resamples <- vfold_cv(train, v=10, strata = BlueTarp)

cv_metrics <- metric_set(accuracy, roc_auc, sens, yardstick::spec, precision, recall, f_meas)

# Create KNN model
tune_nn_spec <- nearest_neighbor(
  engine = "kknn",
  mode = "classification",
  neighbors = tune()
)


# Create workflow
tune_nn_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(tune_nn_spec)

# Tuning range
tune_nn_params <- extract_parameter_set_dials(tune_nn_wf) |>
  update(neighbors = neighbors(c(2,40)))

# Control
cv_control <- control_grid(save_pred = TRUE)


# Tune using cross-validation
tune_results_nn <- tune_grid(
  tune_nn_wf,
  resamples = resamples,
  control = cv_control,
  metrics = cv_metrics,
  grid = grid_regular(tune_nn_params, levels = 39)
)

# Choose best K
show_best(tune_results_nn, metric = "roc_auc", n = 5)

best_nn <- select_best(tune_results_nn, metric = "roc_auc")
```


```{r, fig.width=9, fig.height=11}
# Plot tuning results

autoplot(tune_results_nn) +
  theme_bw() +
  theme(
    text = element_text(size = 13)
  )
```



Result interpretation: The number of neighbors (K) is a tunable hyperparameter in KNN. We evaluated values of K from 2 to 40 using 10-fold cross-validation. This tuning range allowed us to identify the value of K that maximized model performance while avoiding both overfitting (very small K) and underfitting (very large K). The optimal value selected was K=25. The 'best_nn' confirmed that the final model used 25 neighbors. 

The KNN results indicate that the RGB (Red, Green, and Blue) pixel intensity values are highly informative for distinguishing Blue Tarp from Non-Tarp pixels. Using 10-fold cross-validation, the model achieved an average ROC AUC of approximately 0.995, indicating excellent discriminative ability. In other words, the model is able to almost perfectly distinguish Blue Tarp pixels from Non-Tarp pixels. The overall accuracy was also very high at approximately 99.7%. However, because the dataset is highly imbalanced, accuracy alone is not an adequate measure of performance. Therefore, we also examined sensitivity (recall), specificity, precision, F1 score, and ROC AUC.

The tuning plots show that as K increases, the predictions become smoother and the model becomes less sensitive, resulting in a slight decrease in recall. We also observed that ROC AUC continued to increase slightly while the F1 score decreased. This behavior is expected because these metrics evaluate different aspects of model performance. ROC AUC measures how well the model ranks observations across all possible classification thresholds, whereas the F1 score evaluates the balance between precision and recall at a single decision threshold. Consequently, the two metrics do not necessarily reach their maximum values at the same value of K. Overall, the KNN model demonstrated outstanding predictive performance and showed that RGB color information alone is highly effective for identifying blue tarp pixels in this dataset.
