---
title: "Project-1_Group5>"
author: "Sienna Lee, Chloe Dapaah, Michael Awad, Bryson Matsuda"
date: "2026-06-08"
output: html_document
---

```{r}
library(future)
plan(multisession, workers = parallel::detectCores(logical = FALSE))
```

# Training Data
## Preprocess the Data

Create binary variable from Class

```{r}
haiti_train <- read.csv('https://gedeck.github.io/DS-6030/projects/data/HaitiPixels.csv')

haiti_train$BlueTarp <- factor(ifelse(haiti_train$Class == 'Blue Tarp','Blue Tarp', 'Non-Tarp'),
  levels = c("Non-Tarp", "Blue Tarp"))
```


```{r}
# Normalize Color variables

# haiti_train$red_scale <- haiti_train$Red/255 
# haiti_train$green_scale <- haiti_train$Green/255 
# haiti_train$blue_scale <- haiti_train$Blue/255
```

## Exploratory Data Analysis

Blue Tarp vs Non-Blue Tarp count in Data

```{r}
table(haiti_train$BlueTarp)
```

Visualization of the number of Blue Tarp vs Non-Blue Tarp

```{r}
library(ggplot2)

haiti_train |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  geom_bar() +
  scale_fill_manual(values = c('Blue Tarp' = 'skyblue', 'Non-Tarp' = 'darkred'))
  
```

Boxplots for binary against colors

```{r}
#| fig.width: 9
#| fig.height: 8
#| fig.align: center
#| out.width: 75%
library(patchwork)

b1 <- ggplot(haiti_train, aes(x = BlueTarp, y = Red)) +
  geom_boxplot()
b2 <- ggplot(haiti_train, aes(x = BlueTarp, y = Green)) +
  geom_boxplot()
b3 <- ggplot(haiti_train, aes(x = BlueTarp, y = Blue)) +
  geom_boxplot()
b1+b2+b3
```

Histograms of color variables

```{r}
#| fig.width: 7
#| fig.height: 3

g1 <-ggplot(haiti_train, aes(Red)) +
  geom_histogram(fill='darkred')
g2 <- ggplot(haiti_train, aes(Green)) +
  geom_histogram(fill='lightgreen')
g3 <- ggplot(haiti_train, aes(Blue)) +
  geom_histogram(fill='skyblue')
g1+g2+g3
```

# Holdout Data
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
```


## Hold-Out Data: Unzip the file
```{r}
#Create a folder
unzip("HoldOutData.zip", exdir = "HoldOut")

#Check files
list.files("HoldOut")
```

## Parse single ENVI ROI file
```{r}
parse_envi_roi <- function(filepath, label) {
  read_table(
    filepath,
    comment    = ";",
    col_names  = c("ID", "X", "Y", "Map_X", "Map_Y",
                   "Lat", "Lon", "Red", "Green", "Blue"),
    col_types  = cols(.default = col_double()),
    show_col_types = FALSE
  ) |>
    select(Red, Green, Blue) |>
    mutate(
      BlueTarp = factor(label, levels = c(0, 1),
                        labels = c("Non-Tarp", "Blue Tarp"))
    )
}


```

## Load all files and combine into one df
```{r}
files <- tribble(
  ~path,                                      ~label,
  "orthovnir057_ROI_NON_Blue_Tarps.txt",       0,
  "orthovnir067_ROI_Blue_Tarps.txt",           1,
  "orthovnir067_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir069_ROI_Blue_Tarps.txt",           1,
  "orthovnir069_ROI_NOT_Blue_Tarps.txt",       0,
  "orthovnir078_ROI_Blue_Tarps.txt",           1,
  "orthovnir078_ROI_NON_Blue_Tarps.txt",       0
)
 
holdout <- pmap_dfr(files, ~ parse_envi_roi(..1, ..2))
cat("Total rows:", scales::comma(nrow(holdout)), "\n\n")
```

## Data quality checks
```{r}
cat("Dimensions \n")
glimpse(holdout)
 
cat("\n Missing values \n")
print(colSums(is.na(holdout)))
 
cat("\n Value ranges (should be 0–255) \n")
holdout |>
  select(Red, Green, Blue) |>
  summary() |>
  print()
 
cat("\n Duplicate rows \n")
n_dupes <- sum(duplicated(holdout |> select(Red, Green, Blue, BlueTarp)))
cat("Duplicates:", scales::comma(n_dupes), "\n")
 
cat("\n Class distribution \n")
class_counts <- holdout |> count(BlueTarp)
print(class_counts)
 
imbalance_ratio <- class_counts$n[class_counts$BlueTarp == "Non-Tarp"] /
                   class_counts$n[class_counts$BlueTarp == "Blue Tarp"]
cat("Imbalance ratio:", round(imbalance_ratio, 0), ":1 (Non-Tarp : Blue Tarp)\n")
```

## Exploratory Data Analysis

```{r}
holdout |>
  ggplot(aes(x = BlueTarp, fill = BlueTarp)) +
  labs(title = "Tarp Distribution",
       y = "Pixel Count") +
  geom_bar() +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue"))
```


```{r}
haiti_holdout <- holdout |>
  pivot_longer(cols = c(Red, Green, Blue),
               names_to  = "Channel",
               values_to = "Intensity") |>
  mutate(Channel = factor(Channel, levels = c("Red", "Green", "Blue")))
glimpse(haiti_holdout)

ggplot(haiti_holdout, aes(x = BlueTarp, y = Intensity, fill = BlueTarp)) +
  geom_boxplot() +
  facet_wrap(~Channel, ncol = 3) +
  scale_fill_manual(values = c("Non-Tarp" = "grey", "Blue Tarp" = "blue")) +
  labs(title = "Channel Intensity by Class",
       y = "Pixel Intensity (0–255)")
```

# Random Forest Model
## create split for tuning
```{r}
set.seed(1)

split <- initial_split(haiti_train, prop = 0.8, strata = BlueTarp)
train_data <- training(split)
test_data <- testing(split)
```

## creating recipe, cross-validation, and defining model
```{r}
rf_recipe <- recipe(BlueTarp ~ Red + Green + Blue,
  data = train_data)

folds <- vfold_cv(train_data, v = 10, strata = BlueTarp)

rf_spec <- rand_forest(mode = "classification", trees = 1000, min_n = tune(), 
  mtry = tune()) |>
    set_engine("ranger", importance = "impurity")
```

## creating workflow
```{r}
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
```

## tuning
```{r}
rf_grid <- grid_regular(mtry(range = c(1,3)),
  min_n(range = c(2, 30)),
  levels = 5)

my_metrics <- metric_set(accuracy, precision, roc_auc, recall, f_meas, spec)
tuned_rf <- tune_grid(rf_wf, resamples = folds,
  grid = rf_grid, metrics = my_metrics)
```

```{r}
autoplot(tuned_rf)
```

Performance is pretty stable for each metric across min_n

## selecting best model and fitting final model
```{r}
best_rf <- select_best(tuned_rf, metric = "f_meas")
final_rf <- finalize_workflow(rf_wf, best_rf)

rf_fit <- fit(final_rf, data = train_data)
```

## evaluate on test
```{r}
rf_pred <- predict(rf_fit, new_data = test_data, type = "prob") |>
  bind_cols(predict(rf_fit, new_data = test_data), test_data)

my_metrics(
  rf_pred,
  truth = BlueTarp,
  estimate = .pred_class,
  `.pred_Blue Tarp`,
  event_level = "second"
)
```

## confusion matrix
```{r}
rf_pred |>
  conf_mat(
    truth = BlueTarp,
    estimate = .pred_class
  )
```

## final holdout evaluation
```{r}
haiti_holdout_wide <-
  haiti_holdout |>
  mutate(pixel = rep(1:(n()/3), each = 3)) |>
  pivot_wider(
    id_cols = c(pixel, BlueTarp),
    names_from = Channel,
    values_from = Intensity
  ) |>
  select(-pixel)
```
```{r}
final_rf_fit <- fit(final_rf, data = haiti_train)

holdout_pred <- predict(final_rf_fit, new_data = haiti_holdout_wide, type = "prob") |>
  bind_cols(predict(final_rf_fit, new_data = haiti_holdout_wide), haiti_holdout_wide)
```


```{r}
my_metrics(
  holdout_pred,
  truth = BlueTarp,
  estimate = .pred_class,
  `.pred_Blue Tarp`,
  event_level = "second"
)
```

```{r}
holdout_pred |>
  conf_mat(
    truth = BlueTarp,
    estimate = .pred_class
  )
```